Parsing Postfix log files

### 2008-07-21 doc/logparser-introduction.tex
Merged 'Other Postfix log parsers reviewed' and 'Previous research in this area' to create a 'State of the Art Review', which is now section 3.
Removed 'Throughput monitor' review.

Improved introduction and summary in 'State of the Art Review'; miscellaneous
related changes.

Moved 'Graphs' section closer to where it will eventually be, the 'Results'
section.

@@

@@

Appendix~\ref{other-parsers} discusses other parsers and why they were
deemed unsuitable for the task, including why they could not be improved or
expanded upon.

@@

Appendix~\ref{graphs} contains graphs illustrating the topics previously
discussed in \sectionref{rule efficiency} (rule efficiency), and explains
the irregularities observed in these graphs.

>>>>

Section~\ref{state of the art review} reviews both the previously published
research in this area and other available Postfix log parsers, discussing
why they were deemed unsuitable for the task, including why they could not
be improved or expanded upon.

@@

Section~\ref{graphs} contains graphs illustrating the topics previously
discussed in \sectionref{rule efficiency} (rule efficiency), and explains
the irregularities observed in these graphs.

@@

@@

### 2008-07-21 doc/logparser-other-parsers-review.tex
Merged 'Other Postfix log parsers reviewed' and 'Previous research in this area' to create a 'State of the Art Review', which is now section 3.
Removed 'Throughput monitor' review.

Improved introduction and summary in 'State of the Art Review'; miscellaneous
related changes.

Moved 'Graphs' section closer to where it will eventually be, the 'Results'
section.

\section{Other Postfix log parsers reviewed}

@@

It is important to compare and contrast newly developed programs,
algorithms and parsers against those already available, to accurately judge
what, if any, improvements are delivered by the newcomers.  There are not
that many previously developed Postfix log parsers, indeed it was quite
difficult to find ten parsers to review for this project, and the
functionality offered ranges from quite basic to much more mature,
depending on the needs of the creator.  The programs are not reviewed from
an independent viewpoint; the objective is to compare and contrast each
program with \parsername{}.  There are some important differences between
\parsername{} and the parsers reviewed here:

@@

\subsection{Throughput monitor}

This utility tracks the number of events which occurred over a particular
time and warns if the frequency of events passes a certain threshold.  It
is designed to provide real time alerts when dictionary attacks, mail loops
or similar problems occur. It does not attempt to extract or save data,
correlate log lines, or any of the more advanced tasks described in this
document, because it is not designed to do so.  The user must construct the
\regexes{} to match significant input lines before it will parse any log
files.  This program was not reviewed in further detail because its aims
are so far removed from the aims of this project.

\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
checked 2008/05/01.)

@@

Please see the previous in-depth discussion of Log Mail Analyser in
\sectionref{prior art}.

@@

\begin{table}[ht]
\begin{tabular}{l|l|l|l|l|l}
    Parser         & Parsed test & Data            & Custom        & Documentation  & Source \\
                   & log files?  & store?          & reports?      & quality?       & code?  \\
    \hline         &             &                 &               &                &        \\
    Pflogsumm      & Yes         & No              & Partial\dag{} & Good           & Yes    \\
    Sawmill        & Yes         & Yes             & Searches      & Reasonable     & No     \\
    Splunk         & Yes         & Yes             & Searches      & Abundant       & No     \\
                   &             &                 & \& reports    & but poor       &        \\
    Isoqlog        & No          & Yes             & No            & No             & Yes    \\
    AWStats        & Partially   & Yes             & Partial\dag{} & Good           & Yes    \\
    Throughput     & N/A         & No              & No            & Brief but good & Yes    \\
    \tab{}monitor  &             &                 &               &                &        \\
    Anteater       & No          & No              & No            & No             & Yes    \\
    YAALA          & No          & Yes\ddag{}      & Searches      & Poor           & Yes    \\
    Lire           & Yes         & Yes             & Yes           & Yes            & Yes    \\
    Logrep         & No          & Yes             & No            & Barely         & Yes    \\
    \LMA{}         & No          & Yes             & No            & No             & Yes    \\ 
    \parsername{}  & Yes         & Yes\dag{}\dag{} & No            & Yes            & Yes    \\
\end{tabular}

@@

\dag{}\dag{}\parsername{} is the only parser with documentation for its
data store.

>>>>

\section{State of the Art Review}

@@

\label{state of the art review}

Ten other parsers, starting with the only previously published research in
this area, have been reviewed as part of the background research for this
project.  None of the parsers reviewed perform the type of advanced parsing
and log correlation required for this project; all bar one are intended to
perform a specific parsing and reporting task, rather than to be a generic
parser, extracting data and leaving generation of reports from the data to
other programs.  Some parsers save the data extracted to a data store, but
the majority discard all data once they have finished running, making
historical analysis impossible.  The other parsers reviewed all produce a
report of varying complexity and detail, whereas the program described here
does not attempt to produce a report at all; that responsibility is
deferred to a separate program, to be developed later.  The parsing
algorithm and program developed for this project are designed to enable
much more detailed log analysis by providing a stable platform for
subsequent programs to develop upon.

XXX EXTEND THIS\@.

\label{prior art}

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in a
form suitable for a systems administrator to search using the myriad of
standard Unix text processing utilities already available, producing both a
\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
support both \CSV{} and Berkeley DB appears to have been a limiting factor:
\CSV{} is a very simple format where a record is stored in a single line,
with fields separated by a comma or other punctuation symbol.  Problems
with \CSV{} files include the difficulty in escaping separators, providing
multiple values for a field (e.g.\ multiple recipients) and adding new
fields.  There is no facility to document the fields or the separator,
unlike \SQL{} databases where every database includes a schema naming the
fields and the type of data they store (integer, text, timestamp, etc.).

Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
unlike \SQL{} databases which store arbitrary tuples; in \LMA{}'s main
table the key is an integer referred to by secondary tables, and the value
used is a \CSV{} line containing all of the data for that row.  The
secondary by-sender, by-recipient, by-date and by-\IP{} tables use the
sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
referring to the main table is the value.  In effect this re-implements
\SQL{} foreign keys, but without the functionality offered by even the most
basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
requires custom code to search on some combination of the above, though the
authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
of the three output formats: it does not provide the functionality of a
basic \SQL{} database, not can it be used with standard Unix text
processing tools.

The data stored is limited to time and date, hostname and \IP{} address of
client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
codes or remote servers\footnote{A single mail may be sent to multiple
remote servers if it was addressed to recipients in different domains, or
Postfix needs to try multiple servers for one or more recipients.} is not
explained in the paper; experimental observation shows that multiple
records are added when there are multiple recipients (sadly the records are
not associated or linked in any way), and presumably the same approach is
taken when there are multiple destination servers.

The schema used with the MySQL database is undocumented, but at least it is
possible to discover the schema with an \SQL{} database, unlike with
Berkeley DB\@; all \SQL{} databases embed the schema into the database and
provide commands for displaying it.  Berkeley DB does not embed a schema,
as there is neither requirement nor benefit; it only provides \textbf{(key,
value)} pairs, so any additional structuring of the data is imposed by the
application, thus it behoves the application to document this structure.

Parsing in \LMA{} requires major changes to the code to parse new log lines
or extract additional data.  It does not appear to deal with any of the
complications discussed in \sectionref{complications} (initial
complications) and \sectionref{additional complications} (additional
complications).  It does not differentiate between different types of
rejections, so it is not suitable for the purposes of this project; the
data about which restriction caused the rejection is discarded, whereas an
important goal of this project is to retain that data to aid optimisation
and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
files generated on Solaris hosts because the fields added automatically
differ from those added on Linux hosts; log files from Solaris hosts (and
possibly other operating systems) thus require preprocessing before parsing
by \LMA{}.  Once the preprocessing has been performed on the
\numberOFlogFILES{} test log files \LMA{} parses the log files without
complaint, though it does produce 32 entries in its output file for every
rejection in the input log file; it also missed some 40\% of delivered
mail.  Once these glaring deficiencies were discovered the author did not
spend more time checking the results.

\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
DAILY-REJECT initially failed with an error message from the Perl
interpreter; after making some corrections to the code it worked, and
produced the \CSV{} lines for the specified day where the \SMTP{} code
signifies a rejection.  All of these reports are trivially simple to
produce from the \CSV{} file using the standard Unix tool \texttt{awk},
e.g.\ the most complicated, DAILY-REJECT, is merely:

\begin{verbatim}
awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
    { print $0; print " "; }' lma_output.txt
\end{verbatim}

Notes about the command above:

    \item It outputs an unnecessary line containing only a single space
        after each matching line, for compatibility with the output of
        DAILY-REJECT\@.

    \item Some records have an \SMTP{} code of \textit{0}, though this is
        not a valid \SMTP{} code; presumably this is due to \LMA{} not
        parsing a line properly.  \LMA{} considers this code to be a
        rejection, so the command above must, too.

    \item In a tiny proportion of records the \SMTP{} code is
        \textit{deferred\/} (again, not a valid code, and most likely due
        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
        so the equivalent \texttt{awk} command must do so, too.

There are a number of differences in the output from DAILY-REJECT and the
\texttt{awk} command:

    \item \LMA{} does not escape or replace the separators used in the
        \CSV{} records when they are present in sender or recipient
        addresses, leading to some mangled records; these records are not
        present in the output from \texttt{awk}.

    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
        produces only one output line, whereas the \texttt{awk} command
        produces all matching lines.

In summary, \LMA{} appears to be a proof of concept, written to demonstrate
the point of their paper (that having this information in an accessible
fashion is useful to systems administrators), rather than a program
designed to be extensible and useful in a production environment.

% Literature review notes:
%
% Hard-coded parsing, requiring code changes to add more.  Attempts to
% correlate log lines, saves data to database for data mining purposes.
% Hard to extend/expand/understand.  Appears to only save: date and hour,
% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
% receiver and e-mail status (sent, rejected).  Undocumented schema.
% Design decision to use \CSV{} as an intermediate format between the log
% file and the database seems to have been restrictive.  Appears to require
% a queueid but majority of log entries (e.g.\ rejections) lack a queueid.
% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
% when generating reports/data mining would be preferable.  Supporting
% Berkeley DB is probably limiting the software - an example is the
% difficulty in searching a pipe-delimited string, so they have
% re-implemented foreign keys with tables keyed by ip address etc.\
% pointing at the main table - this also will not scale well.  There does
% not appear to be any attempt to deal with the complications I have
% encountered: their parsing is not detailed enough to encounter them.  It
% does not run properly; does not create any output; throws up errors.

It is important to compare and contrast newly developed programs,
algorithms and parsers against those already available, to accurately judge
what, if any, improvements are delivered by the newcomers.  There are not
that many previously developed Postfix log parsers --- indeed it was quite
difficult to find ten parsers to review for this project --- and the
functionality offered ranges from quite basic to much more mature,
depending on the needs of the creator.  The programs are not reviewed from
an independent viewpoint; the objective is to compare and contrast each
program with \parsername{}.  There are some important differences between
\parsername{} and the parsers reviewed here:

@@

@@

@@

\begin{table}[ht]
\begin{tabular}{llllll}
    Parser         & Parsed test & Data            & Custom        & Documentation  & Source \\
                   & log files?  & store?          & reports?      & quality?       & code?  \\
    \tableline{}
    \LMA{}         & No          & Yes             & No            & Poor           & Yes    \\ 
    Pflogsumm      & Yes         & No              & Partial\dag{} & Good           & Yes    \\
    Sawmill        & Yes         & Yes             & Searches      & Reasonable     & No     \\
    Splunk         & Yes         & Yes             & Searches      & Abundant       & No     \\
                   &             &                 & \& reports    & but poor       &        \\
    Isoqlog        & No          & Yes             & No            & Poor           & Yes    \\
    AWStats        & Partially   & Yes             & Partial\dag{} & Good           & Yes    \\
    Anteater       & No          & No              & No            & Poor           & Yes    \\
    YAALA          & No          & Yes\ddag{}      & Searches      & Poor           & Yes    \\
    Lire           & Yes         & Yes             & Yes           & Reasonable     & Yes    \\
    Logrep         & No          & Yes             & No            & Poor           & Yes    \\
    \parsername{}  & Yes         & Yes\nialpha{}   & \nibeta{}     &                & Yes    \\
    \tableline{}
\end{tabular}

@@

\nialpha{} \parsername{} is the only parser with documentation for its data
store.

\nibeta{} \parsername{} defers report generation to subsequent programs,
but all the necessary data and documentation to produce reports is
provided.
